---
title: Collecting Highway Bearings
author: D. Leedal
date: '2020-07-06'
slug: collecting-highway-bearings
categories:
  - R
tags:
  - R
  - geospatial
  - sf
  - OSM
subtitle: ''
summary: ''
authors: []
lastmod: '2020-07-06T11:35:31+01:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

### What direction is this road going?

I'm kicking off this blog with a geospatial example. I did some work recently where I needed to know:

* the nearest section of road to a given point; and,
* the road's approximate bearing.

It's a long story why this was needed but getting there involved some useful discoveries so I'm repeating the crux of it here. Interestingly, shortly after writing the scripts, I came across [Geoff Boeing's beautiful post](https://geoffboeing.com/2018/07/comparing-city-street-orientations/) visualising similar information for 25 cities. 

The best approach I could think of was to use OpenStreetMap data. Firstly associate a point with it's nearest highway 'way' feature. Then get the coordinates of the beginning and end of the feature and use the pair of points to define a bearing. Makes sense? OK, here we go:

### Get the packages
```{r}
library(geosphere)
library(sf)
library(leaflet)
library(osmdata)
library(data.table)
library(ggmap)
```

### Organise the data
To keep things manageable I'm going to use 4 points dotted around the Northern Quarter in Manchester (UK). I'll put them in a data.frame:

```{r}
df <- data.frame(ID = 1:4,
 Longitude = c(-2.232086, -2.235675, -2.233628, -2.232453),
 Latitude = c(53.484510, 53.484318, 53.484103, 53.481176), stringsAsFactors = FALSE)
```

Then I'll convert to an `sf` (spatial features) multi point object. I use this command over and over.

```{r}
df.sf <- st_as_sf(df, coords = c("Longitude", "Latitude"), crs=4326)
```

The crs argument is telling st_as_sf that these coordinates are WGS84 by referencing the appropriate EPSG code 4326 (it's a good idea to memorise that number!). This is needed later when I project the points out of latitude/longitude format and into distance-from-bottom-left format. If you're not familiar with this concept [here's a good pdf tutorial](https://www.nceas.ucsb.edu/sites/default/files/2020-04/OverviewCoordinateReferenceSystems.pdf).

### Visualise
At this point I'll take a look at the 4 points. I'm using leaflet which works well with Blogdown.
```{r}
m <- leaflet(df) %>% addTiles() %>% addCircleMarkers(
  lng = ~Longitude,
  lat = ~Latitude,
    radius = 8,
    color = "green",
    stroke = FALSE, fillOpacity = 0.9
  )
m
```

### Get information from OpenStreetMap
The `opq` query from the `osmdata` package forms an OSM overpass filter and sends it to OSM's servers. The result is an `osmdata` object containing various items. The item I'm interested in is the `osm_lines` which is a set of `sf` linestrings. These are the way objects representing highway segments. The query has returned 44 highway segments that intersected the bounded box of our 4 points (at the time of posting).

```{r}
this.bbox <- st_bbox(df.sf)
osm.road.types = c("motorway", "motorway_link", "trunk", "trunk_link", "primary", "primary_link", "secondary", 
                   "secondary_link", "tertiary", "unclassified", "residential")
out <- opq (this.bbox) %>%
            #add_osm_feature (key = "highway", value = c(osm.road.types, paste0(osm.road.types, "_link"))) %>%
            osmdata::add_osm_feature (key = "highway", value = osm.road.types) %>%
            osmdata::osmdata_sf(quiet = F)
out
road.lines <- out$osm_lines
nrow(road.lines)
```

### Get candidate closest roads to points
Soon I'll find the closest highway segment to each point, but if I had thousands of points and highways it would be wise to only check a subset of highways by associating nearby highways to each point. I'll do that next. First I'll project the points and the road segments to a suitable coordinate reference system (for Manchester, the UK Ordnance Survay EPSG: 27700 is a good choice), apply a 50m buffer to the points, then identify which road segments intersect with this buffer.

```{r}
df.sf.tf <- st_transform(df.sf, 27700) # US National Atlas Equal Area projection
road.lines.tf <- st_transform(road.lines, 27700)
buffer.size <- 50
df.sf.tf.bf <- st_buffer(df.sf.tf, buffer.size)
intersected.items <- st_intersects(df.sf.tf.bf, road.lines.tf, sparse = F)
dim(intersected.items)
```

The dimensions show I have 4 rows, one for each of the 4 points, and 44 columns, one for each OSM road segment within the bounding box. The matrix contains TRUE/FALSE values where TRUE indicates a point of intersection.

### Get the closest road segment
So now, using the location of TRUE values in each row or the `intersected.items` matrix, I can subset the `road.lines` so that the nearest highway search is constrained to a much smaller number of highway segments (ones within 50m).
The `st_coordinates` function is useful to return just the coordinate information from an `sf` object, and the `dist2Line` function from the `geosphere` package calculates the shortest distance between a point and a polyline (There are some good details about doing this type of calculation at the [Movable Type website](https://www.movable-type.co.uk/scripts/latlong.html).

```{r}
number.of.point <- nrow(intersected.items)
distances.vect <- matrix(NA_real_, nrow = nrow(df.sf.tf), ncol = 4)
colnames(distances.vect) <- c("distance", "lon", "lat", "ID")
for (i in seq_len(number.of.point)) {
  # Out of all 44 road segments, get the index of the ones that are within 50m of point i. 
  original.road.idx <- which(intersected.items[i, ] == TRUE)
  distances.vect.temp  <- dist2Line(p = st_coordinates(df.sf[i, ]), line = as_Spatial(road.lines[original.road.idx, ]))
  distances.vect.temp[1, "ID"] <- original.road.idx[distances.vect.temp[1, "ID"]]
  distances.vect[i, ] <- distances.vect.temp
}
distances.vect
```

So now I've got the nearest road segments *and* the coordinates of the nearest point on the road segments to the original points. These are stored in the lon/lat columns of `distances.vec`. So the original vague points are now *snapped* to actual highway locations which is a bonus.

### Get the road section bearings
Next I'll extract the chosen 4 road segments from the full set of 44 and add the useful features I've just created so far. Getting the coordinates of the first and last points is a bit fiddly. I'm using `data.table` for this. When retrieving the coordinates via `st_coordinates` a third column labelled `L1` is created with a group ID of each polyline object. This can be used as a grouping variable in data.table. The `.SD[...]` syntax holds the sub-data.table once the full data.table has been split by the grouping variable, finally 1 and `.N` are the first row and last row of each sub-data.table. There's other ways to do this but the speed of `data.table` would pay off if I had thousands of points. 
Once we have the beginning and end coordinate pairs it's straightforward to call the `bearing` function from the `geosphere` package. 
```{r}
out <- road.lines[as.integer(distances.vect[, "ID"]), ]
out$dist.to.nearest <- distances.vect[, "distance"]
out$snap.lat <- distances.vect[, "lat"]
out$snap.lon <- distances.vect[, "lon"]
matching.road.lines.coords <- data.table::as.data.table(st_coordinates(out))
matching.road.lines.coords.begin <- as.matrix(matching.road.lines.coords[, .SD[1, .(X, Y)], by = "L1"][, L1 := NULL])
matching.road.lines.coords.end <- as.matrix(matching.road.lines.coords[, .SD[.N, .(X, Y)], by = "L1"][, L1 := NULL])
matching.road.lines.bearing <- geosphere::bearing(matching.road.lines.coords.begin, matching.road.lines.coords.end)
out$bearing <- (matching.road.lines.bearing + 360) %% 360
out$bearing
```

### Plot with estimated bearing
In order to visualise the estimated road bearing for my 4 points I need to add arrow endpoints information. Some trigonometry wil accomplish this (the 0.00025 value is determines the length of the direction arrow):
```{r}
out$delta.x <- sin((pi*out$bearing) / 180) * 0.00025
out$delta.y <- cos((pi*out$bearing) / 180) * 0.00025
```

Now I can plot the snapped locations and show an arrow indicating the highway bearing over a basemap for visual validation.

```{r echo = FALSE, message = FALSE, warning = FALSE}
n.quarter.bb <- c(left = -2.236, bottom = 53.48, right = -2.23, top = 53.485)
n.quarter <- get_stamenmap(bbox = n.quarter.bb, zoom = 18, messaging = FALSE)
ggmap(n.quarter) + 
  geom_point(data = out, aes(x=snap.lon, y=snap.lat), colour = "darkgreen", size = 4) +
  geom_segment(data = out, aes(x=snap.lon, y=snap.lat, xend = snap.lon + delta.x, yend = snap.lat + delta.y), colour = "darkgreen", arrow = arrow(length = unit(0.2,"cm")))
```

### Summary
The results are pretty good, definitely good enough to classify a point into one of the 8 compass bearing directions: N, NE, E... etc. In this post I've explored working with `sf` spatial points, generated and run an OSM overpass filter, and done some useful visualisation with `Leaflet` and `ggmap`. I'm sure there are things I could tidy and optimise here but I hope you found this a little useful.



